---
title: "Gene Ontology (GO) enrichment analysis using clusterProfiler per cell type for sc-RNA-Seq Analysis in 10X Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  OrgDb_value: '.'
  resolution_list: '.'
  n_value: '.'
  root_dir: './'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```

\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak
  
# Information about this notebook

After completing clustering and cell type annotation in the scRNA‑seq workflow, a common next step is to functionally interpret the resulting gene sets. Gene Ontology (GO) enrichment enables biological insight by identifying over‑represented biological processes, molecular functions, or cellular components among differentially expressed genes (DEGs).

In our sc/snRNA‑seq pipeline, GO analysis is incorporated as part of downstream functional interpretation following DE analysis and identification of gene sets via methods such as [FindAllMarkers()](https://satijalab.org/seurat/reference/findallmarkers) or [ClosestFeature()](https://stuartlab.org/signac/reference/closestfeature). 

We will use clusterProfiler R package for GO enrichment because it supports multiple organisms, accepts various gene ID formats (SYMBOL, Ensembl), and integrates seamlessly with Seurat and Bioconductor annotation databases.

# Set up

```{r load-library, echo=TRUE}
attach(params)
suppressPackageStartupMessages({
  library(tidyverse)
  #library(Seurat)
  library(knitr)
  library(DT)
  library(glue)

  library(htmlwidgets)
  library(clusterProfiler)
  library(enrichplot)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
}) 
```

# Directories and paths to file Inputs/Outputs

```{r set-dir-and-file-names, echo=TRUE}
analysis_dir <- file.path(root_dir, "analyses", "de-go-analysis") 
module_results_dir <- file.path(analysis_dir, "results")
module_plots_dir <- file.path(analysis_dir, "plots")
markers_dir_module <- file.path(analysis_dir, "results", "01-de-volcano-plots")

# Input files
markers_file <- file.path(markers_dir_module, glue::glue("Res_{resolution_list}_Markers_all.tsv"))

# Create plots directory
go_plots_dir <- file.path(module_plots_dir, "02-go-clusterprofiler") 
if (!dir.exists(go_plots_dir)) {
  dir.create(go_plots_dir)}

# Source functions
source(paste0(analysis_dir, "/util/function-go-clusterprofiler-markers.R"))
```

```{r echo=FALSE, warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H')
a4width <- 8.3
a4height <- 11.7
```


# Read Markers File per Cell Type

```{r read-markers-cell-type, echo=TRUE}
all.markers <- read.csv(markers_file, sep = "\t", header = TRUE)
```

```{r markers-cell-type-interactive-table, echo=TRUE}
# Interactive Table
datatable(all.markers, 
          options = list(pageLength = 5, 
                         autoWidth = TRUE, 
                         server = TRUE), 
          filter = "top")
```

We will filter and keep `r n_value` markers for visualization purposes.

```{r find-markers-cell-type-top-n_value, echo=TRUE}
top_n_value <- all.markers %>% 
  group_by(cluster) %>% 
  top_n(n = n_value, wt = avg_log2FC)

write.table(top_n_value, file = paste(results_dir, "/", "Res_", resolution_list, glue::glue("_Markers_top_{n_value}.tsv"), sep = ""), sep = "\t", quote = FALSE, row.names = FALSE)
```

# Gene Ontology (GO) enrichment analysis using clusterProfiler

In this step, we conduct Gene Ontology (GO) enrichment analysis to identify biological processes associated with cluster-specific accessible regions (e.g., differential expression). Using the [clusterProfiler](https://bioconductor.org/packages/devel/bioc/html/clusterProfiler.html) package, we assess whether any GO terms are significantly enriched among these gene sets. This approach helps interpret the functional implications of differential expression across clusters, i.e., predicted cell type.

```{r go-clusterProfiler, fig.width = 15, fig.height = 8, fig.fullwidth = TRUE, echo=TRUE}
ego_results <- run_clusterwise_GO_enrichment_markers(closest_genes_all = top_n_value, 
                                                     OrgDb_value = OrgDb_value, 
                                                     plots_dir = go_plots_dir)
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r echo=FALSE}
sessionInfo()
```

